Spatial phylogenetics of Japanese ferns: Patterns, processes, and implications for conservation

### 2021-09-01 ms/manuscript.Rmd
Add deer damage analysis to MS

```{r load-data}
# Load objects from targets workflow
tar_load(c(
  "aic_env_repro",
  "binary_sig_results",
  "biodiv_ferns_cent",
  "biodiv_ferns_spatial",
  "broad_alignment_list",
  "comm_ferns",
  "comm_ferns_full",
  "csl_file",
  "fern_traits",
  "green_list",
  "japan_fern_tree",
  "japan_map_points",
  "japan_pterido_tree",
  "japan_rbcL",
  "japan_shp",
  "k_phylogeny",
  "k_taxonomy",
  "lat_span_summary",
  "lrt_results",
  "model_fits",
  "model_params",
  "ms_functions",
  "occ_point_data_summary",
  "phy_sig_results",
  "plastome_calibration_dates",
  "plastome_tree",
  "ppgi",
  "protected_areas",
  "refs_other_yaml",
  "refs_yaml",
  "repro_data",
  "signif_cells_protected_area",
  "template_file",
  "traits_summary"
))
```

@@

**Methods** We used data from >`r occ_sum$filtered$n_specimens$value %>% plyr::round_any(1000, f = floor) %>% number(big.mark = ",")` fern specimens in Japan to compile a community dataset of `r nrow(comm_ferns) %>% scales::number(big.mark = ",")` 0.2° grid-cells including `r ncol(comm_ferns)` native, non-hybrid taxa. 
We combined this with a phylogenetic tree and functional trait data to analyze taxonomic, phylogenetic, and functional diversity, and modeled the distribution of biodiversity in response to environmental factors and reproductive mode. 
We conducted categorical analysis of neo- and paleo-endemism to identify areas of significant phylogenetic endemism. 
We used hierarchical clustering of taxonomic and phylogenetic distances to analyze bioregions. 
We compared the overlap of significantly diverse grid-cells with existing conservation zones to assess conservation status.

**Results** We observe three major patterns in the spatial distribution of biodiversity: that of taxonomic richness, which is highest at mid-latitudes, that of phylogenetic and functional diversity and phylogenetic endemism, which are highest in small southern islands, and that of relative phylogenetic and functional diversity, which are high at both high and low latitudes, and low at mid-latitudes. 
Grid-cells were grouped in to three (phylogenetic) or four (taxonomic) major bioregions. 
Temperature and apomixis were identified as likely drivers of patterns of taxonomic and phylogenetic diversity.
Conservation status (% protected area) is generally higher than the overall rate for grid-cells with significantly high biodiversity.

**Main conclusions** The integrative approach used here reveals previously undetected patterns and drivers of biodiversity in the ferns of Japan. 
The different distributions of each diversity type reflect the major bioregions. 
Although peak areas of different diversity types are largely non-overlapping, each one has a relatively high protection status. 
Despite this, future conservation efforts must be cognizant of dangers not addressed here such as climate change and herbivory by deer.

<!-- may include title words -->
**Key words:** apomixis, biodiversity, biogeography, CANAPE, ferns, Japan, phylogenetic diversity.

@@

### Assessment of conservation status

We downloaded SHP files of protected areas in Japan from the Japan Ministry of the Environment (https://www.biodic.go.jp/biodiversity/activity/policy/map/map17/index.html). 
We only considered areas that include protection of plants, and categorized protected areas as either "high" (no human activities allowed at all) or "medium" (some economic activities allowed by permit) following [@Kusumoto2017]. 
We cropped grid-cells to only include land area, then calculated the percent of protected area of each status type within grid-cells with high levels of biodiversity as measured with PD, FD, PE, or taxonomic richness. 
For PD, FD, and PE, significance was assessed with a one-tailed test at alpha = 0.05 against the 999 null communities described above. 
For taxonomic richness, cells ranked in the top >5% were considered highly diverse.

@@

### Conservation status

@@

# Extract values for medium and high protection level
total_percent_protected_medium <- total_percent_protected %>% filter(status == "medium") %>% pull(percent_protected)
total_percent_protected_high <- total_percent_protected %>% filter(status == "high") %>% pull(percent_protected)
total_percent_protected_all <- sum(total_percent_protected_medium, total_percent_protected_high)
```

The percentage of protected area in grid-cells with significantly high biodiversity range from `r min_protected_total` to `r max_protected_total` total, including `r min_protected_high` to `r max_protected_high` with high status and `r min_protected_med` to `r max_protected_med` with medium status (`r figure("conserv-status")`). 
Areas with high PE were the best protected (`r best_protected_high` high, `r best_protected_med` medium, `r best_protected_total` total). 
Cells with significantly high biodiversity generally had a greater percentage of area protected than the overall rate for Japan (`r scales::percent(total_percent_protected_high, accuracy = 0.1)` high, `r scales::percent(total_percent_protected_medium, accuracy = 0.1)` medium, `r scales::percent(total_percent_protected_all, accuracy = 0.1)` total); only FD was lower than the overall rate for highly protected areas.

@@

### Conservation status

Although the different facets of biodiversity have different distribution patterns, we found that grid-cells with significantly high amounts of biodiversity tend to be relatively well protected compared to the baseline protection rate, regardless of biodiversity metric (`r figure("conserv-status")`). 
One possible reason is that in the past, some ferns have been included in lists of rare and endemic plants for consideration in designation of conservation areas (A. Ebihara, pers. comm.). 
Another (as suggested previously) is that ferns may serve as bioindicators reflecting overall levels of biodiversity, particularly for other plant groups.

Despite the good news that ferns enjoy relatively strong protection status in Japan, we must bear in mind that there are other threats posed that conservation zones cannot ameliorate, the most obvious being climate change. 
One other such threat is herbivory by native Japanese deer, *Cervus nippon*. 
Populations of these deer have increased rapidly since the 1970s due to a combination of climate change and a decrease in human hunters and natural predators [@Takatsuki2009]. 
At dense populations, the deer can prevent forest regeneration and result in plant communities dominated by a limited number of species unpalatable to deer [@Takatsuki2009], although lower population sizes may promote plant diversity as per the intermediate disturbance hypothesis [@Nishizawa2016].
They have caused extensive damage even within national parks [@Tokita2006], which have the highest protection status. 
Although various efforts to prevent deer herbivory have been deployed such as culling and exclusion fences [@Takatsuki2009], much work remains to be done to save vulnerable plant populations.

@@

```{r conserv-status-make}
# Plot protected areas
conserv_status_plot <- signif_cells_protected_area %>%
  mutate(
    metric = case_when(
      metric == "pd" ~ "PD",
      metric == "fd" ~ "FD",
      metric == "pe" ~ "PE",
      metric == "richness" ~ "Richness",
    ),
    status = str_to_sentence(status)) %>%
  mutate(metric = factor(metric, levels = c("PE", "FD", "PD", "Richness"))) %>%
  ggplot(aes(x = percent_protected, y = metric, fill = status)) +
  geom_col(position = position_dodge()) +
  geom_vline(xintercept = total_percent_protected_medium, color = "grey40", linetype = 3) +
  geom_vline(xintercept = total_percent_protected_high, color = "grey40", linetype = 2) +
  scale_fill_manual(values = protection_cols) +
  scale_x_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  labs(
    x = "Percent protected",
    y = "Biodiversity metric",
    fill = "Protection status"
  ) +
  theme_bw() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "bottom"
  )
```

```{r conserv-status-save, eval = params$doc_type == "doc"}
# Save plot if rendering to doc
ggsave(result_file(figure("conserv-status"), "pdf"), conserv_status_plot, width = 4, height = 4, units = "in")
```

```{r conserv-status-print, fig.width = 4, fig.height = 4, eval = params$doc_type == "pdf"}
# Print plot if rendering to PDF
conserv_status_plot
```

**`r figure("conserv-status")`**. Percent of land area with protected status for grid cells with significantly high biodiversity by at least one metric. Biodiversity metrics include phylogenetic diversity (PD), functional diversity (FD), and phylogenetic endemism (PE). Significance of biodiversity metrics assessed by a one-tailed test comparing observed values to a null distribution of 999 random values (see Materials and Methods). Vertical lines indicate total percent area protected across Japan regardless of degree of biodiversity; dashed, high protection level (`r scales::percent(total_percent_protected_high, accuracy = 0.1)`); dotted, medium protection level (`r scales::percent(total_percent_protected_medium, accuracy = 0.1)`).

>>>>

```{r load-data}
# Load objects from targets workflow
tar_load(c(
  "aic_env_repro",
  "binary_sig_results",
  "biodiv_ferns_cent",
  "biodiv_ferns_spatial",
  "broad_alignment_list",
  "comm_ferns",
  "comm_ferns_full",
  "csl_file",
  "deer_range",
  "fern_traits",
  "green_list",
  "japan_fern_tree",
  "japan_map_points",
  "japan_pterido_tree",
  "japan_rbcL",
  "japan_shp",
  "k_phylogeny",
  "k_taxonomy",
  "lat_span_summary",
  "lrt_results",
  "model_fits",
  "model_params",
  "ms_functions",
  "occ_point_data_summary",
  "phy_sig_results",
  "plastome_calibration_dates",
  "plastome_tree",
  "ppgi",
  "protected_areas",
  "refs_other_yaml",
  "refs_yaml",
  "repro_data",
  "signif_cells_protected_area",
  "signif_cells_deer_danger",
  "template_file",
  "traits_summary"
))
```

@@

**Methods** We used data from >`r occ_sum$filtered$n_specimens$value %>% plyr::round_any(1000, f = floor) %>% number(big.mark = ",")` fern specimens in Japan to compile a community dataset of `r nrow(comm_ferns) %>% scales::number(big.mark = ",")` 0.2° grid-cells including `r ncol(comm_ferns)` native, non-hybrid taxa. 
We combined this with a phylogenetic tree and functional trait data to analyze taxonomic, phylogenetic, and functional diversity, and modeled the distribution of biodiversity in response to environmental factors and reproductive mode. 
We conducted categorical analysis of neo- and paleo-endemism to identify areas of significant phylogenetic endemism. 
We used hierarchical clustering of taxonomic and phylogenetic distances to analyze bioregions. 
We compared the overlap of significantly diverse grid-cells with existing conservation zones and range maps of native Japanese deer to assess conservation status and threats.

**Results** We observe three major patterns in the spatial distribution of biodiversity: that of taxonomic richness, which is highest at mid-latitudes, that of phylogenetic and functional diversity and phylogenetic endemism, which are highest in small southern islands, and that of relative phylogenetic and functional diversity, which are high at both high and low latitudes, and low at mid-latitudes. 
Grid-cells were grouped in to three (phylogenetic) or four (taxonomic) major bioregions. 
Temperature and apomixis were identified as likely drivers of patterns of taxonomic and phylogenetic diversity.
Conservation status (% protected area) was generally high for grid-cells with significantly high biodiversity across all biodiversity metrics, but the threat due to herbivory by native Japanese deer varied by biodiversity metric.

**Main conclusions** The integrative approach used here reveals previously undetected patterns and drivers of biodiversity in the ferns of Japan. 
The different distributions of each diversity type reflect the major bioregions. 
Although peak areas of different diversity types are largely non-overlapping, each one has relatively high protection coverage at medium protection level. 
Despite this, future conservation efforts must recognize that threats can vary by biodiversity metric, and consider threats not directly analyzed here such as climate change.

<!-- may include title words -->
**Key words:** apomixis, biodiversity, biogeography, conservation, CANAPE, ferns, Japan, phylogenetic diversity.

@@

### Assessment of conservation status and threats

We downloaded SHP files of protected areas in Japan from the Japan Ministry of the Environment (https://www.biodic.go.jp/biodiversity/activity/policy/map/map17/index.html) and Ministry of Land, Infrastructure, Transport and Tourism (https://nlftp.mlit.go.jp/ksj/gml/datalist/KsjTmplt-A45.html). 
We only considered areas that include protection of plants, and categorized protected areas as either "high" (no human activities allowed at all) or "medium" status (some economic activities allowed by permit) [@Kusumoto2017]. 
Any areas that overlapped within a protection status were combined, and areas that overlapped between medium and high status were considered only high status.
We downloaded distribution maps (SHP and TIF) of native Japanese deer (*Cervus nippon*), an herbivore that has caused extensive damage to native plant communities due to rapid population growth since the 1970s [@Takatsuki2009] and poses a high risk to ferns, from the Japan Ministry of the Environment (https://www.biodic.go.jp/biodiversity/activity/policy/map/map14/index.html).
To assess change in threat due to herbivory by deer over time, we used three maps available in this dataset: range of Japanese deer surveyed in 1978, range surveyed in 2003, and estimated range inferred from a model including snow cover and forest type based on the 2003 survey data [@kankyosho2021].
We cropped grid-cells to only include land area, then calculated the percent of protected area of each status type or percent of area occupied by deer for each survey period within grid-cells with high levels of biodiversity as measured with PD, FD, PE, or taxonomic richness. 
For PD, FD, and PE, significance was assessed with a one-tailed test at alpha = 0.05 against the 999 null communities described above. 
For taxonomic richness, cells ranked in the top >5% were considered highly diverse.

@@

### Conservation status and threats

@@

# Extract values for medium and high protection level
total_percent_protected_medium <- total_percent_protected %>% filter(status == "medium") %>% pull(percent_protected)
total_percent_protected_high <- total_percent_protected %>% filter(status == "high") %>% pull(percent_protected)
total_percent_protected_all <- sum(total_percent_protected_medium, total_percent_protected_high)

### Deer ###

# Calculate baseline protection percentage: percent protected areas across the whole country
total_percent_deer_danger <-
  deer_range %>%
  mutate(area = st_area(.) %>% units::set_units(km^2)) %>%
  st_drop_geometry %>%
  as_tibble() %>%
  group_by(range) %>%
  summarize(
    total_area = sum(area),
    .groups = "drop"
  ) %>%
  mutate(percent_danger = total_area/area_of_japan) %>%
  mutate(percent_danger = as.numeric(percent_danger))

# Convert to a list for easy indexing
deer_overall <- total_percent_deer_danger %>%
  mutate(
    # Format numbers for nice printing
    area = total_area %>% as.numeric %>% scales::number(accuracy = 1, big.mark = ",") %>% paste("km^2^"),
    percent = scales::percent(percent_danger, accuracy = 0.1)) %>%
  split(.$range)

deer_in_high_biodiv <-
signif_cells_deer_danger %>%
  group_by(range) %>%
  summarize(percent_danger = mean(percent_danger), .groups = "drop") %>%
  mutate(
    # Format numbers for nice printing
    percent = scales::percent(percent_danger, accuracy = 0.1)) %>%
  split(.$range)

deer_highest_danger <- signif_cells_deer_danger %>%
  slice_max(percent_danger) %>%
  # Make sure biodiv type with greatest danger is richness in estimated dataset
  verify(metric == "richness") %>%
  verify(range == "estimated") %>%
  mutate(
    # Format numbers for nice printing
    percent = scales::percent(percent_danger, accuracy = 0.1)) %>%
  split(.$range)
```

The percentage of protected area in grid-cells with significantly high biodiversity range from `r min_protected_total` to `r max_protected_total` total, including `r min_protected_high` to `r max_protected_high` with high status and `r min_protected_med` to `r max_protected_med` with medium status (`r figure("conserv-status")`a). 
Cells with significantly high biodiversity had a greater percentage of area protected than the overall rate for Japan at the medium protection status (`r scales::percent(total_percent_protected_medium, accuracy = 0.1)`) across all biodiversity metrics.
In contrast, cells with significantly high biodiversity had a lower percentage of area protected than the overall rate for Japan at the high protection status (`r scales::percent(total_percent_protected_medium, accuracy = 0.1)`) for most biodiversity metrics.
Only cells with significantly high PE had a higher percentage area protected (`r best_protected_high` high, `r best_protected_med` medium, `r best_protected_total` total) than the overall rate at both levels of protection.

The overall range of Japanese deer in Japan has increased from `r deer_overall[["1978"]]$area` (`r deer_overall[["1978"]]$percent`) in 1978 to `r deer_overall[["2003"]]$area` (`r deer_overall[["2003"]]$percent`) in 2003, and may be as high as `r deer_overall$estimated$area` (`r deer_overall$estimated$percent`) according to model estimates.
The percentage of area occupied by deer in grid-cells with significantly high biodiversity increased from `r deer_in_high_biodiv[["1978"]]$percent` in 1978 to `r deer_in_high_biodiv[["2003"]]$percent` in 2003 when averaged across biodiversity types.
Cells with high FD, PD, or PE tended to have less deer occupation compared to the overall rate (`r figure("conserv-status")`b).
However, cells with high species richness consistently exceeded the overall rate of deer presence, with up to `r deer_highest_danger$estimated$percent` occupied by deer in the estimated range dataset.

@@

### Conservation status and threats

We found that grid-cells with significantly high amounts of biodiversity are better protected than the baseline protection rate at the medium protection level across all biodiversity metrics, and better than the baseline rate at the high protection level for PE (`r figure("conserv-status")`a). 
One possible reason for this generally high level of protection is that in the past, some ferns have been included in lists of rare and endemic plants for consideration in designation of conservation areas (A. Ebihara, pers. comm.). 
In particular, there has been an emphasis on conserving rare and endemic species in Japanese southern islands such as Ogasawara and Amami (FIXME: add ref), which include many grid-cells with high PE.
Another reason (as suggested previously) is that ferns may serve as bioindicators reflecting overall levels of biodiversity, particularly for other plant groups.

Despite the good news that ferns enjoy relatively strong protection status in Japan, we must bear in mind that there are other threats posed that conservation zones cannot ameliorate, the most obvious being climate change. 
One other such threat is herbivory by native Japanese deer, *Cervus nippon*. 
The rapid increase in population size observed in surveys from 1978 to 2003 is likely due to a combination of climate change and a decrease in human hunters and natural predators [@Takatsuki2009]. 
At dense populations, the deer can prevent forest regeneration and result in plant communities dominated by a limited number of species unpalatable to deer [@Takatsuki2009], although lower population sizes may promote plant diversity as per the intermediate disturbance hypothesis [@Nishizawa2016].
They have caused extensive damage even within national parks [@Tokita2006], which have the highest protection status. 
Although various efforts to prevent deer herbivory have been deployed such as culling and exclusion fences [@Takatsuki2009], much work remains to be done to save vulnerable plant populations.

Our analysis shows that the threat to fern biodiversity posed by Japanese deer differs strongly by biodiversity type; although grid-cells with significantly high FD, PD, and PE tend to have lower rates of deer occupancy compared to the baseline rate, grid-cells with high richness experience much greater co-occurrence with deer than the baseline rate (`r figure("conserv-status")`b).
This pattern is due to the fact that Japanese deer do not occur in the southern islands, which house high amounts of PD and PE; rather, areas with high species richness are primarily located in southern Honshu, where deer expansion has been intense (FIXME: add SI fig).
This result highlights the importance of using multiple metrics of biodiversity for establishing conservation priorities.

@@

```{r conserv-status-make}
# Plot protected areas
a <- signif_cells_protected_area %>%
  mutate(
    metric = case_when(
      metric == "pd" ~ "PD",
      metric == "fd" ~ "FD",
      metric == "pe" ~ "PE",
      metric == "richness" ~ "Richness",
    ),
    status = str_to_sentence(status)) %>%
  mutate(metric = factor(metric, levels = c("PE", "FD", "PD", "Richness"))) %>%
  ggplot(aes(x = percent_protected, y = metric, fill = status)) +
  geom_col(position = position_dodge()) +
  geom_vline(xintercept = total_percent_protected_medium, color = "grey40", linetype = 3) +
  geom_vline(xintercept = total_percent_protected_high, color = "grey40", linetype = 2) +
  scale_fill_manual(
    values = protection_cols,
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      label.position = "left"
    )) +
  scale_x_continuous(
    labels = function(x) scales::percent(x, accuracy = 1),
    expand = expansion(mult = c(0, .05))) +
  labs(
    x = "Percent protected",
    y = "Biodiversity metric",
    fill = "Protection status"
  ) +
  theme_bw() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "bottom"
  )

# Plot areas in danger from herbivory by deer 
b <- signif_cells_deer_danger %>%
  mutate(
    metric = case_when(
      metric == "pd" ~ "PD",
      metric == "fd" ~ "FD",
      metric == "pe" ~ "PE",
      metric == "richness" ~ "Richness",
    ),
    range = str_to_sentence(range)) %>%
  mutate(metric = factor(metric, levels = c("PE", "FD", "PD", "Richness"))) %>%
  ggplot(aes(x = percent_danger, y = metric, fill = range)) +
  geom_col(position = position_dodge()) +
  # line for 1978 overall percent
  geom_vline(
    xintercept = total_percent_deer_danger %>% filter(range == "1978") %>% pull(percent_danger), 
    color = "grey40", linetype = 3) +
  # line for 2003 overall percent
  geom_vline(
    xintercept = total_percent_deer_danger %>% filter(range == "2003") %>% pull(percent_danger), 
    color = "grey40", linetype = 2) +
  # line for estimated overall percent
  geom_vline(
    xintercept = total_percent_deer_danger %>% filter(range == "estimated") %>% pull(percent_danger), 
    color = "grey40", linetype = 1) +
  scale_fill_manual(
    values = deer_colors,
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      label.position = "left"
    )) +
  scale_x_continuous(
    labels = function(x) scales::percent(x, accuracy = 1),
    expand = expansion(mult = c(0, .05))) +
  labs(
    x = "Percent area with deer",
    y = "Biodiversity metric",
    fill = "Range data source"
  ) +
  theme_bw() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "bottom"
  )

conserv_status_plot <- a + b +
  plot_layout(ncol = 2) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  ) &
  theme(plot.tag = element_text(face = "bold"))
```

```{r conserv-status-save, eval = params$doc_type == "doc"}
# Save plot if rendering to doc
ggsave(result_file(figure("conserv-status"), "pdf"), conserv_status_plot, width = 7, height = 4, units = "in")
```

```{r conserv-status-print, fig.width = 7, fig.height = 4, eval = params$doc_type == "pdf"}
# Print plot if rendering to PDF
conserv_status_plot
```

**`r figure("conserv-status")`**. Conservation status and threat due to herbivory by deer for ferns of Japan by biodiversity metric. (a) Percent of land area with protected status for grid cells with significantly high biodiversity. (b) Percent of land area occupied by deer for grid cells with significantly high biodiversity. Biodiversity metrics include phylogenetic diversity (PD), functional diversity (FD), and phylogenetic endemism (PE). Significance of biodiversity metrics assessed by a one-tailed test comparing observed values to a null distribution of 999 random values (see Materials and Methods). For (a), vertical lines indicate total percent area protected across Japan (baseline protection rate); dashed, high protection level (`r scales::percent(total_percent_protected_high, accuracy = 0.1)`); dotted, medium protection level (`r scales::percent(total_percent_protected_medium, accuracy = 0.1)`). For (b), vertical lines indicate total percentage of land area occupied by deer across Japan (baseline threat rate); solid, percentage estimated with a model based on 2003 survey data (`r deer_overall$estimated$percent`); dashed, percentage based on 2003 survey data (`r deer_overall[["2003"]]$percent`); dotted, percentage based on 1978 survey data (`r deer_overall[["1978"]]$percent`).
